# Read the Data into R
```{r}
library(survey)
library(skimr)
data <- readr::read_csv("yrs24.csv", guess_max=400000)
skimr::skim(data)
data$record
```
#Filtering Data by Year to understand non-missing values
```{R}
library(dplyr)
library(skimr)
data.91 <- select(data, year, PSU, 23:102)
data_1991 <- data.91 %>%
install.packages("survey")
library(survey)
library(skimr)
data <- readr::read_csv("yrs24.csv", guess_max=400000)
skimr::skim(data)           
```
# Filtering Data by Year to understand non-missing values
```{R}
library(dplyr)
library(skimr)
pca.data <- select(data, year, PSU, 23:102)
data_1991 <- pca.data %>%
  filter(year== "1991")
skimr::skim(data_1991)
```
```{R}
library(dplyr)
library(skimr)
data.92 <- select(data, year, PSU, 23:102)
data_1992 <- data.92 %>%
pca.data <- select(data, year, PSU, 23:102)
data_1992 <- pca.data %>%
  filter(year== "1992")
skimr::skim(data_1992)
```

```{R}
library(dplyr)
library(skimr)
data.93 <- select(data, year, PSU, 23:102)
data_1993 <- data.93 %>%
pca.data <- select(data, year, PSU, 23:102)
data_1993 <- pca.data %>%
  filter(year== "1993")
skimr::skim(data_1993)
```

```{R}
library(dplyr)
library(skimr)
data.94 <- select(data, year, PSU, 23:102)
data_1994 <- data.94 %>%
pca.data <- select(data, year, PSU, 23:102)
data_1994 <- pca.data %>%
filter(year== "1994")
skimr::skim(data_1994)
```

```{R}
library(dplyr)
library(skimr)
data.95 <- select(data, year, PSU, 23:102)
data_1995 <- data.95 %>%
  filter(year== "2019")
skimr::skim(data_1995)
```
#It appears as though odd years are answered for the questionnaire 
```{R}
library(dplyr)
library(skimr)
slim.data <- select(data, year, c(23,24,27,28,30,31,32,33,41,42,43,44,45,47,
                                                        48,55,56,60,61,62,65,66,70,71,72,73,74,75,
                                                        76,77,78,79,80,81,82,83,96))

pca.data <- filter(slim.data, year %in% c(1991,1993, 1995, 1997, 1999, 2001, 2003,
                                          2005, 2007, 2009, 2011, 2013, 2015, 2017,
                                          2019))

pca.data <- subset(pca.data, select = -c(year))
skimr::skim(pca.data)
```
#PCA of Items to Slim Down Variables
```{r}
library(psych)
fit <- principal(pca.data, nfactors=17, rotate="oblimin")
summary(fit)
fit
```

#Now that we know componenets, need to recode variables, and build components
```{r}
pca.data$q29new[pca.data$q29==1] <- 1
pca.data$q29new[pca.data$q29==2] <- 3
pca.data$q29new[pca.data$q29==3] <- 2

pca.data$q46new[pca.data$q46==1] <- 1
pca.data$q46new[pca.data$q46==2] <- 7
pca.data$q46new[pca.data$q46==3] <- 6
pca.data$q46new[pca.data$q46==4] <- 5
pca.data$q46new[pca.data$q46==5] <- 4
pca.data$q46new[pca.data$q46==6] <- 3
pca.data$q46new[pca.data$q46==7] <- 2

pca.data$q58new[pca.data$q58==1] <- 2
pca.data$q58new[pca.data$q58==2] <- 1

pca.data$q59new[pca.data$q59==1] <- 1
pca.data$q59new[pca.data$q59==2] <- 8
pca.data$q59new[pca.data$q59==3] <- 7
pca.data$q59new[pca.data$q59==4] <- 6
pca.data$q59new[pca.data$q59==5] <- 5
pca.data$q59new[pca.data$q59==6] <- 4
pca.data$q59new[pca.data$q59==7] <- 3
pca.data$q59new[pca.data$q59==8] <- 2

pca.data$q62new[pca.data$q62==1] <- 1
pca.data$q62new[pca.data$q62==2] <- 3
pca.data$q62new[pca.data$q62==3] <- 2

pca.data$q64new[pca.data$q64==1] <- 1
pca.data$q64new[pca.data$q64==6] <- 2
pca.data$q64new[pca.data$q64==5] <- 3
pca.data$q64new[pca.data$q64==3] <- 4
pca.data$q64new[pca.data$q64==4] <- 5
pca.data$q64new[pca.data$q64==7] <- 6
pca.data$q64new[pca.data$q64==8] <- 7
pca.data$q64new[pca.data$q64==2] <- 8

pca.data$Sexual_Activity_Behavior = pca.data$q58new + pca.data$q59new + pca.data$q60 + pca.data$q61 + pca.data$q62 + pca.data$q63 + pca.data$q64new

pca.data$Marijuana_Use_Behavior = pca.data$q45 +  pca.data$q46new +  pca.data$q47

pca.data$Cigarette_Smoking_Behavior = pca.data$q32 + pca.data$q33

pca.data$Other_Drug_Use = pca.data$q50 + pca.data$q51 + pca.data$q55 + pca.data$q56

pca.data$Suicide_Attempt_Behavior = pca.data$q28 + pca.data$q29new

pca.data$Weapon_Carrying_Behavior = pca.data$q12 + pca.data$q13

pca.data
```
#Now, create the original data set, with the new variables and componenets included
```{R}
yrisk <- merge(data,pca.data, all.x=TRUE)
yrisk
```