# Read the Data into R
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(skimr)
```

```{r}
library(survey)
library(skimr)
data <- readr::read_csv("yrs24.csv", guess_max=400000)
skimr::skim(data)
data$record
```
#Filtering Data by Year to understand non-missing values
```{R}
library(dplyr)
library(skimr)
data.91 <- select(data, year, PSU, 23:102)
data_1991 <- data.91 %>%
install.packages("survey")
library(survey)
library(skimr)
data <- readr::read_csv("yrs24.csv", guess_max=400000)
skimr::skim(data)           
```
# Filtering Data by Year to understand non-missing values
```{R}
library(dplyr)
library(skimr)
pca.data <- select(data, year, PSU, 23:102)
data_1991 <- pca.data %>%
  filter(year== "1991")
skimr::skim(data_1991)
```
```{R}
library(dplyr)
library(skimr)
data.92 <- select(data, year, PSU, 23:102)
data_1992 <- data.92 %>%
pca.data <- select(data, year, PSU, 23:102)
data_1992 <- pca.data %>%
  filter(year== "1992")
skimr::skim(data_1992)
```

```{R}
library(dplyr)
library(skimr)
data.93 <- select(data, year, PSU, 23:102)
data_1993 <- data.93 %>%
pca.data <- select(data, year, PSU, 23:102)
data_1993 <- pca.data %>%
  filter(year== "1993")
skimr::skim(data_1993)
```

```{R}
library(dplyr)
library(skimr)
data.94 <- select(data, year, PSU, 23:102)
data_1994 <- data.94 %>%
pca.data <- select(data, year, PSU, 23:102)
data_1994 <- pca.data %>%
filter(year== "1994")
skimr::skim(data_1994)
```

```{R}
library(dplyr)
library(skimr)
data.95 <- select(data, year, PSU, 23:102)
data_1995 <- data.95 %>%
  filter(year== "2019")
skimr::skim(data_1995)
```
#It appears as though odd years are answered for the questionnaire 
```{R}
library(dplyr)
library(skimr)
slim.data <- select(data, year, c(23,24,27,28,30,31,32,33,41,42,43,44,45,47,
                                                        48,55,56,60,61,62,65,66,70,71,72,73,74,75,
                                                        76,77,78,79,80,81,82,83,96))

pca.data <- filter(slim.data, year %in% c(1991,1993, 1995, 1997, 1999, 2001, 2003,
                                          2005, 2007, 2009, 2011, 2013, 2015, 2017,
                                          2019))

pca.data <- subset(pca.data, select = -c(year))
skimr::skim(pca.data)
```
#PCA of Items to Slim Down Variables
```{r}
library(psych)
fit <- principal(pca.data, nfactors=17, rotate="oblimin")
summary(fit)
fit
```

#Now that we know componenets, need to recode variables, and build components
```{r}
pca.data$q29new[pca.data$q29==1] <- 1
pca.data$q29new[pca.data$q29==2] <- 3
pca.data$q29new[pca.data$q29==3] <- 2

pca.data$q46new[pca.data$q46==1] <- 1
pca.data$q46new[pca.data$q46==2] <- 7
pca.data$q46new[pca.data$q46==3] <- 6
pca.data$q46new[pca.data$q46==4] <- 5
pca.data$q46new[pca.data$q46==5] <- 4
pca.data$q46new[pca.data$q46==6] <- 3
pca.data$q46new[pca.data$q46==7] <- 2

pca.data$q58new[pca.data$q58==1] <- 2
pca.data$q58new[pca.data$q58==2] <- 1

pca.data$q59new[pca.data$q59==1] <- 1
pca.data$q59new[pca.data$q59==2] <- 8
pca.data$q59new[pca.data$q59==3] <- 7
pca.data$q59new[pca.data$q59==4] <- 6
pca.data$q59new[pca.data$q59==5] <- 5
pca.data$q59new[pca.data$q59==6] <- 4
pca.data$q59new[pca.data$q59==7] <- 3
pca.data$q59new[pca.data$q59==8] <- 2

pca.data$q62new[pca.data$q62==1] <- 1
pca.data$q62new[pca.data$q62==2] <- 3
pca.data$q62new[pca.data$q62==3] <- 2

pca.data$q64new[pca.data$q64==1] <- 1
pca.data$q64new[pca.data$q64==6] <- 2
pca.data$q64new[pca.data$q64==5] <- 3
pca.data$q64new[pca.data$q64==3] <- 4
pca.data$q64new[pca.data$q64==4] <- 5
pca.data$q64new[pca.data$q64==7] <- 6
pca.data$q64new[pca.data$q64==8] <- 7
pca.data$q64new[pca.data$q64==2] <- 8

pca.data$Sexual_Activity_Behavior = pca.data$q58new + pca.data$q59new + pca.data$q60 + pca.data$q61 + pca.data$q62 + pca.data$q63 + pca.data$q64new

pca.data$Marijuana_Use_Behavior = pca.data$q45 +  pca.data$q46new +  pca.data$q47

pca.data$Cigarette_Smoking_Behavior = pca.data$q32 + pca.data$q33

pca.data$Other_Drug_Use = pca.data$q50 + pca.data$q51 + pca.data$q55 + pca.data$q56

pca.data$Suicide_Attempt_Behavior = pca.data$q28 + pca.data$q29new

pca.data$Weapon_Carrying_Behavior = pca.data$q12 + pca.data$q13

pca.data
```
#Now, create the original data set, with the new variables and componenets included
```{R}
yrisk <- merge(data,pca.data, all.x=TRUE)
yrisk
```

#Summarize Data by Question and Year
```{r}
gyear <- yrisk %>%
  mutate(year) %>%
  select(year, age, sex, grade, stheight, stweight, bmi, bmipct, q8, q9, q12, q13, q15, q16, q17, q18, q26, q27, q28, q29, q30, q32, q33, q36, q40, q41, q45, q46, q47, q50, q51, q55, q56, q57, q58, q59, q60, q61, q62, q63, q64, q67, q68, q69, q70, q83)
```

```{r}
grouped_gyear <- gyear %>%
  group_by(year)
grouped_gyear
```

# Means of questions identified by PCA by year. 
```{r}
myearq8 <- grouped_gyear %>%
  select(year, q8) %>%
  group_by(year) %>%
  drop_na(q8) %>%
  summarise(avg = mean(q8))
myearq8
```

```{r}
myearq9 <- grouped_gyear %>%
  select(year, q9) %>%
  group_by(year) %>%
  drop_na(q9) %>%
  summarise(avg = mean(q9))
myearq9
```

```{r}
myearq12 <- grouped_gyear %>%
  select(year, q12) %>%
  group_by(year) %>%
  drop_na(q12) %>%
  summarise(avg = mean(q12))
myearq12
```

```{r}
myearq13 <- grouped_gyear %>%
  select(year, q13) %>%
  group_by(year) %>%
  drop_na(q13) %>%
  summarise(avg = mean(q13))
myearq13
```

```{r}
myearq15 <- grouped_gyear %>%
  select(year, q15) %>%
  group_by(year) %>%
  drop_na(q15) %>%
  summarise(avg = mean(q15))
myearq15
```

```{r}
myearq16 <- grouped_gyear %>%
  select(year, q16) %>%
  group_by(year) %>%
  drop_na(q16) %>%
  summarise(avg = mean(q16))
myearq16
```

```{r}
myearq17 <- grouped_gyear %>%
  select(year, q17) %>%
  group_by(year) %>%
  drop_na(q17) %>%
  summarise(avg = mean(q17))
myearq17
```

```{r}
myearq18 <- grouped_gyear %>%
  select(year, q18) %>%
  group_by(year) %>%
  drop_na(q18) %>%
  summarise(avg = mean(q18))
myearq18
```

```{r}
myearq26 <- grouped_gyear %>%
  select(year, q26) %>%
  group_by(year) %>%
  drop_na(q26) %>%
  summarise(avg = mean(q26))
myearq26
```

```{r}
myearq27 <- grouped_gyear %>%
  select(year, q27) %>%
  group_by(year) %>%
  drop_na(q27) %>%
  summarise(avg = mean(q27))
myearq27
```

```{r}
myearq28 <- grouped_gyear %>%
  select(year, q28) %>%
  group_by(year) %>%
  drop_na(q28) %>%
  summarise(avg = mean(q28))
myearq28
```

```{r}
myearq29 <- grouped_gyear %>%
  select(year, q29) %>%
  group_by(year) %>%
  drop_na(q29) %>%
  summarise(avg = mean(q29))
myearq29
```

```{r}
myearq30 <- grouped_gyear %>%
  select(year, q30) %>%
  group_by(year) %>%
  drop_na(q30) %>%
  summarise(avg = mean(q30))
myearq30
```

```{r}
myearq32 <- grouped_gyear %>%
  select(year, q32) %>%
  group_by(year) %>%
  drop_na(q32) %>%
  summarise(avg = mean(q32))
myearq32
```

```{r}
myearq33 <- grouped_gyear %>%
  select(year, q33) %>%
  group_by(year) %>%
  drop_na(q33) %>%
  summarise(avg = mean(q33))
myearq33
```

```{r}
myearq40 <- grouped_gyear %>%
  select(year, q40) %>%
  group_by(year) %>%
  drop_na(q40) %>%
  summarise(avg = mean(q40))
myearq40
```

```{r}
myearq41 <- grouped_gyear %>%
  select(year, q41) %>%
  group_by(year) %>%
  drop_na(q41) %>%
  summarise(avg = mean(q41))
myearq41
```

```{r}
myearq45 <- grouped_gyear %>%
  select(year, q45) %>%
  group_by(year) %>%
  drop_na(q45) %>%
  summarise(avg = mean(q45))
myearq45
```

```{r}
myearq46 <- grouped_gyear %>%
  select(year, q46) %>%
  group_by(year) %>%
  drop_na(q46) %>%
  summarise(avg = mean(q46))
myearq46
```

```{r}
myearq47 <- grouped_gyear %>%
  select(year, q47) %>%
  group_by(year) %>%
  drop_na(q47) %>%
  summarise(avg = mean(q47))
myearq47
```

```{r}
myearq50 <- grouped_gyear %>%
  select(year, q50) %>%
  group_by(year) %>%
  drop_na(q50) %>%
  summarise(avg = mean(q50))
myearq50
```

```{r}
myearq51 <- grouped_gyear %>%
  select(year, q51) %>%
  group_by(year) %>%
  drop_na(q51) %>%
  summarise(avg = mean(q51))
myearq51
```

```{r}
myearq55 <- grouped_gyear %>%
  select(year, q55) %>%
  group_by(year) %>%
  drop_na(q55) %>%
  summarise(avg = mean(q55))
myearq55
```

```{r}
myearq56 <- grouped_gyear %>%
  select(year, q56) %>%
  group_by(year) %>%
  drop_na(q56) %>%
  summarise(avg = mean(q56))
myearq56
```

```{r}
myearq57 <- grouped_gyear %>%
  select(year, q57) %>%
  group_by(year) %>%
  drop_na(q57) %>%
  summarise(avg = mean(q57))
myearq57
```

```{r}
myearq58 <- grouped_gyear %>%
  select(year, q58) %>%
  group_by(year) %>%
  drop_na(q58) %>%
  summarise(avg = mean(q58))
myearq58
```

```{r}
myearq59 <- grouped_gyear %>%
  select(year, q59) %>%
  group_by(year) %>%
  drop_na(q59) %>%
  summarise(avg = mean(q59))
myearq59
```

```{r}
myearq60 <- grouped_gyear %>%
  select(year, q60) %>%
  group_by(year) %>%
  drop_na(q60) %>%
  summarise(avg = mean(q60))
myearq60
```

```{r}
myearq61 <- grouped_gyear %>%
  select(year, q61) %>%
  group_by(year) %>%
  drop_na(q61) %>%
  summarise(avg = mean(q61))
myearq61
```

```{r}
myearq62 <- grouped_gyear %>%
  select(year, q62) %>%
  group_by(year) %>%
  drop_na(q62) %>%
  summarise(avg = mean(q62))
myearq62
```

```{r}
myearq63 <- grouped_gyear %>%
  select(year, q63) %>%
  group_by(year) %>%
  drop_na(q63) %>%
  summarise(avg = mean(q63))
myearq63
```

```{r}
myearq64 <- grouped_gyear %>%
  select(year, q64) %>%
  group_by(year) %>%
  drop_na(q64) %>%
  summarise(avg = mean(q64))
myearq64
```

```{r}
myearq67 <- grouped_gyear %>%
  select(year, q67) %>%
  group_by(year) %>%
  drop_na(q67) %>%
  summarise(avg = mean(q67))
myearq67
```

```{r}
myearq68 <- grouped_gyear %>%
  select(year, q68) %>%
  group_by(year) %>%
  drop_na(q68) %>%
  summarise(avg = mean(q68))
myearq68
```

```{r}
myearq69 <- grouped_gyear %>%
  select(year, q69) %>%
  group_by(year) %>%
  drop_na(q69) %>%
  summarise(avg = mean(q69))
myearq69
```

```{r}
myearq70 <- grouped_gyear %>%
  select(year, q70) %>%
  group_by(year) %>%
  drop_na(q70) %>%
  summarise(avg = mean(q70))
myearq70
```

```{r}
myearq83 <- grouped_gyear %>%
  select(year, q83) %>%
  group_by(year) %>%
  drop_na(q83) %>%
  summarise(avg = mean(q83))
myearq83
```

# Graph seat belt use over time
---
```{r}
# grouped_gyear
gq8 <- filter(grouped_gyear, !is.na(q8)) %>%
  ggplot + 
  geom_bar(mapping = aes(x = year, fill = factor(q8)), position = "fill", stat = "count", na.rm = FALSE) + 
  labs(title = "How often do you wear a seatbelt when riding in a car driven by someone else?", x = "Year") +
  scale_x_continuous(breaks = c(1991, 1993, 1995, 1997, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 2017, 2019)) +
  scale_fill_manual("Seatbelt Use", values = c('#a6cee3', '#1f78b4', '#fc8d62', '#7570b3', '#d95f02'), labels = c("Never", "Rarely", "Sometimes", "Most", "Always")) + 
theme_classic()
gq8
```
The greatest variability over time is increase in portion of youth always wearing a seatbelt.
---



#Why the NA in later years?
```{r}
grouped_gyear %>%
  dplyr::group_by(year) %>%
  skim()
```
#Number of respondents by year
```{r}
num_peryear <- grouped_gyear %>%
  summarize(n=n())
num_peryear
```

#Graph of Respondents
---
```{r}
gres <- ggplot(num_peryear, aes(x = year, y = n)) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  labs(title = "Number of Respondents per Year", x = "Year", y = "Number") +
  scale_x_continuous(breaks = c(1991, 1993, 1995, 1997, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 2017, 2019))
gres
```
 There was a large increase in number of respondents from 1999 to 2015. 
 The number of respondents decreased sharply from 2015 to 2017. 
 ---
  

# Need to Recode Answers to q40 to create graph
# Graph age of first drink over time
---
```{r}
grouped_gyear$q40new[grouped_gyear$q40==1] <- 1
grouped_gyear$q40new[grouped_gyear$q40==2] <- 7
grouped_gyear$q40new[grouped_gyear$q40==3] <- 6
grouped_gyear$q40new[grouped_gyear$q40==4] <- 5
grouped_gyear$q40new[grouped_gyear$q40==5] <- 4
grouped_gyear$q40new[grouped_gyear$q40==6] <- 3
grouped_gyear$q40new[grouped_gyear$q40==7] <- 2
grouped_gyear

gq40 <-filter(grouped_gyear, !is.na(q40)) %>%
  ggplot + 
  geom_bar(mapping = aes(x = year, fill = factor(q40)), position = "fill", stat = "count", na.rm = TRUE) + 
  labs(title = "At what age did you first have more than a few sips of alcohol?", x = "Year") +
  scale_x_continuous(breaks = c(1991, 1993, 1995, 1997, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 2017, 2019)) +
  scale_fill_manual("First Alcohol", values = c('#a6cee3', '#1f78b4', '#fc8d62', '#7570b3', '#d95f02', "#660066", "#101010"), labels = c("Never", "17 or older", "15 or 16", "13 or 14", "11 or 12", "9 or 10", "8 or younger")) +
  theme_classic()
gq40
```
Overall, youth are older before having more than a few sips of alcohol.
---

# Q30 Looks interesting
---
```{r}
gq30 <- filter(grouped_gyear, !is.na(q30)) %>%
  ggplot + 
  geom_bar(mapping = aes(x = year, fill = factor(q30)), position = "fill", stat = "count", na.rm = FALSE) + 
  labs(title = "Have you ever tried a cigarette, even one or two puffs?", x = "Year") +
  scale_x_continuous(breaks = c(1991, 1993, 1995, 1997, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 2017, 2019)) +
  scale_fill_manual("Tried Cigarettes", values = c('#660066', '#1f78b4'), na.translate = FALSE, labels = c("Yes", "No")) + 
  theme_classic()
gq30
```
Trying cigarettes has gone down over time. 
Our hypothesis is that it is related to an increase in vaping.
But there is not enough data on vaping to draw any conclusions.
---

#Q58
---
```{r}
gq58 <- filter(grouped_gyear, !is.na(q58))  %>%
  ggplot + 
  geom_bar(mapping = aes(x = year, fill = factor(q58)), position = "fill", stat = "count", na.rm = TRUE) + 
  labs(title = "Have you ever had sexual intercourse?", x = "Year") +
  scale_x_continuous(breaks = c(1991, 1993, 1995, 1997, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 2017, 2019)) +
  scale_fill_manual("Had Sex", values = c('#660066', '#1f78b4'), labels = c("Yes", "No")) + 
  theme_classic()
gq58
```
Portion of youth engaging in sexual intercourse has really not changed.
---
