# Read the Data into R
```{r}
library(survey)
library(skimr)
data <- readr::read_csv("yrs24.csv", guess_max=400000)
skimr::skim(data)
data$record
```
#Filtering Data by Year to understand non-missing values
```{R}
library(dplyr)
library(skimr)
data.91 <- select(data, year, PSU, 23:102)
data_1991 <- data.91 %>%
install.packages("survey")
library(survey)
library(skimr)
data <- readr::read_csv("yrs24.csv", guess_max=400000)
skimr::skim(data)           
```
# Filtering Data by Year to understand non-missing values
```{R}
library(dplyr)
library(skimr)
pca.data <- select(data, year, PSU, 23:102)
data_1991 <- pca.data %>%
  filter(year== "1991")
skimr::skim(data_1991)
```
```{R}
library(dplyr)
library(skimr)
data.92 <- select(data, year, PSU, 23:102)
data_1992 <- data.92 %>%
pca.data <- select(data, year, PSU, 23:102)
data_1992 <- pca.data %>%
  filter(year== "1992")
skimr::skim(data_1992)
```

```{R}
library(dplyr)
library(skimr)
data.93 <- select(data, year, PSU, 23:102)
data_1993 <- data.93 %>%
pca.data <- select(data, year, PSU, 23:102)
data_1993 <- pca.data %>%
  filter(year== "1993")
skimr::skim(data_1993)
```

```{R}
library(dplyr)
library(skimr)
data.94 <- select(data, year, PSU, 23:102)
data_1994 <- data.94 %>%
pca.data <- select(data, year, PSU, 23:102)
data_1994 <- pca.data %>%
filter(year== "1994")
skimr::skim(data_1994)
```

```{R}
library(dplyr)
library(skimr)
data.95 <- select(data, year, PSU, 23:102)
data_1995 <- data.95 %>%
  filter(year== "2019")
skimr::skim(data_1995)
```
#It appears as though odd years are answered for the questionnaire 
```{R}
library(dplyr)
library(skimr)
slim.data <- select(data, year, c(23,24,27,28,30,31,32,33,41,42,43,44,45,47,
                                                        48,55,56,60,61,62,65,66,70,71,72,73,74,75,
                                                        76,77,78,79,80,81,82,83,96))

pca.data <- filter(slim.data, year %in% c(1991,1993, 1995, 1997, 1999, 2001, 2003,
                                          2005, 2007, 2009, 2011, 2013, 2015, 2017,
                                          2019))

pca.data <- subset(pca.data, select = -c(year))
skimr::skim(pca.data)
```
#PCA of Items to Slim Down Variables
```{r}
library(psych)
fit <- principal(pca.data, nfactors=17, rotate="oblimin")
summary(fit)
fit
```

#Now that we know componenets, need to recode variables, and build components
```{r}
pca.data$q29new[pca.data$q29==1] <- 1
pca.data$q29new[pca.data$q29==2] <- 3
pca.data$q29new[pca.data$q29==3] <- 2

pca.data$q46new[pca.data$q46==1] <- 1
pca.data$q46new[pca.data$q46==2] <- 7
pca.data$q46new[pca.data$q46==3] <- 6
pca.data$q46new[pca.data$q46==4] <- 5
pca.data$q46new[pca.data$q46==5] <- 4
pca.data$q46new[pca.data$q46==6] <- 3
pca.data$q46new[pca.data$q46==7] <- 2

pca.data$q58new[pca.data$q58==1] <- 2
pca.data$q58new[pca.data$q58==2] <- 1

pca.data$q59new[pca.data$q59==1] <- 1
pca.data$q59new[pca.data$q59==2] <- 8
pca.data$q59new[pca.data$q59==3] <- 7
pca.data$q59new[pca.data$q59==4] <- 6
pca.data$q59new[pca.data$q59==5] <- 5
pca.data$q59new[pca.data$q59==6] <- 4
pca.data$q59new[pca.data$q59==7] <- 3
pca.data$q59new[pca.data$q59==8] <- 2

pca.data$q62new[pca.data$q62==1] <- 1
pca.data$q62new[pca.data$q62==2] <- 3
pca.data$q62new[pca.data$q62==3] <- 2

pca.data$q64new[pca.data$q64==1] <- 1
pca.data$q64new[pca.data$q64==6] <- 2
pca.data$q64new[pca.data$q64==5] <- 3
pca.data$q64new[pca.data$q64==3] <- 4
pca.data$q64new[pca.data$q64==4] <- 5
pca.data$q64new[pca.data$q64==7] <- 6
pca.data$q64new[pca.data$q64==8] <- 7
pca.data$q64new[pca.data$q64==2] <- 8

pca.data$Sexual_Activity_Behavior = pca.data$q58new + pca.data$q59new + pca.data$q60 + pca.data$q61 + pca.data$q62 + pca.data$q63 + pca.data$q64new

pca.data$Marijuana_Use_Behavior = pca.data$q45 +  pca.data$q46new +  pca.data$q47

pca.data$Cigarette_Smoking_Behavior = pca.data$q32 + pca.data$q33

pca.data$Other_Drug_Use = pca.data$q50 + pca.data$q51 + pca.data$q55 + pca.data$q56

pca.data$Suicide_Attempt_Behavior = pca.data$q28 + pca.data$q29new

pca.data$Weapon_Carrying_Behavior = pca.data$q12 + pca.data$q13

pca.data
```
#Now, create the original data set, with the new variables and componenets included
```{R}
yrisk <- merge(data,pca.data, all.x=TRUE)
yrisk
```

#Summarize Data by Question and Year
```{r}
library(dplyr)
gyear <- yrisk %>%
  mutate(year) %>%
  select(year, age, sex, grade, stheight, stweight, bmi, bmipct, q8, q9, q12, q13, q15, q16, q17, q18, q26, q27, q28, q29, q30, q32, q33, q40, q41, q45, q46, q47, q50, q51, q55, q56, q57, q58, q59, q60, q61, q62, q63, q64, q67, q68, q69, q70, q83)
```

```{r}
grouped_gyear <- gyear %>%
  group_by(year)
grouped_gyear
```

# Means of questions identified by PCA by year. 
```{r}
library(tidyverse)

myearq8 <- grouped_gyear %>%
  select(year, q8) %>%
  group_by(year) %>%
  drop_na(q8) %>%
  summarise(avg = mean(q8))
myearq8
```

```{r}
library(tidyverse)

myearq9 <- grouped_gyear %>%
  select(year, q9) %>%
  group_by(year) %>%
  drop_na(q9) %>%
  summarise(avg = mean(q9))
myearq9
```

```{r}
library(tidyverse)

myearq12 <- grouped_gyear %>%
  select(year, q12) %>%
  group_by(year) %>%
  drop_na(q12) %>%
  summarise(avg = mean(q12))
myearq12
```

```{r}
library(tidyverse)

myearq13 <- grouped_gyear %>%
  select(year, q13) %>%
  group_by(year) %>%
  drop_na(q13) %>%
  summarise(avg = mean(q13))
myearq13
```

```{r}
library(tidyverse)

myearq15 <- grouped_gyear %>%
  select(year, q15) %>%
  group_by(year) %>%
  drop_na(q15) %>%
  summarise(avg = mean(q15))
myearq15
```

```{r}
library(tidyverse)

myearq16 <- grouped_gyear %>%
  select(year, q16) %>%
  group_by(year) %>%
  drop_na(q16) %>%
  summarise(avg = mean(q16))
myearq16
```

```{r}
library(tidyverse)

myearq17 <- grouped_gyear %>%
  select(year, q17) %>%
  group_by(year) %>%
  drop_na(q17) %>%
  summarise(avg = mean(q17))
myearq17
```

```{r}
library(tidyverse)

myearq18 <- grouped_gyear %>%
  select(year, q18) %>%
  group_by(year) %>%
  drop_na(q18) %>%
  summarise(avg = mean(q18))
myearq18
```

```{r}
library(tidyverse)

myearq26 <- grouped_gyear %>%
  select(year, q26) %>%
  group_by(year) %>%
  drop_na(q26) %>%
  summarise(avg = mean(q26))
myearq26
```

```{r}
library(tidyverse)

myearq27 <- grouped_gyear %>%
  select(year, q27) %>%
  group_by(year) %>%
  drop_na(q27) %>%
  summarise(avg = mean(q27))
myearq27
```

```{r}
library(tidyverse)

myearq28 <- grouped_gyear %>%
  select(year, q28) %>%
  group_by(year) %>%
  drop_na(q28) %>%
  summarise(avg = mean(q28))
myearq28
```

```{r}
library(tidyverse)

myearq29 <- grouped_gyear %>%
  select(year, q29) %>%
  group_by(year) %>%
  drop_na(q29) %>%
  summarise(avg = mean(q29))
myearq29
```

```{r}
library(tidyverse)

myearq30 <- grouped_gyear %>%
  select(year, q30) %>%
  group_by(year) %>%
  drop_na(q30) %>%
  summarise(avg = mean(q30))
myearq30
```

```{r}
library(tidyverse)

myearq32 <- grouped_gyear %>%
  select(year, q32) %>%
  group_by(year) %>%
  drop_na(q32) %>%
  summarise(avg = mean(q32))
myearq32
```

```{r}
library(tidyverse)

myearq33 <- grouped_gyear %>%
  select(year, q33) %>%
  group_by(year) %>%
  drop_na(q33) %>%
  summarise(avg = mean(q33))
myearq33
```

```{r}
library(tidyverse)

myearq40 <- grouped_gyear %>%
  select(year, q40) %>%
  group_by(year) %>%
  drop_na(q40) %>%
  summarise(avg = mean(q40))
myearq40
```

```{r}
library(tidyverse)

myearq41 <- grouped_gyear %>%
  select(year, q41) %>%
  group_by(year) %>%
  drop_na(q41) %>%
  summarise(avg = mean(q41))
myearq41
```

```{r}
library(tidyverse)

myearq45 <- grouped_gyear %>%
  select(year, q45) %>%
  group_by(year) %>%
  drop_na(q45) %>%
  summarise(avg = mean(q45))
myearq45
```

```{r}
library(tidyverse)

myearq46 <- grouped_gyear %>%
  select(year, q46) %>%
  group_by(year) %>%
  drop_na(q46) %>%
  summarise(avg = mean(q46))
myearq46
```

```{r}
library(tidyverse)

myearq47 <- grouped_gyear %>%
  select(year, q47) %>%
  group_by(year) %>%
  drop_na(q47) %>%
  summarise(avg = mean(q47))
myearq47
```

```{r}
library(tidyverse)

myearq50 <- grouped_gyear %>%
  select(year, q50) %>%
  group_by(year) %>%
  drop_na(q50) %>%
  summarise(avg = mean(q50))
myearq50
```

```{r}
library(tidyverse)

myearq51 <- grouped_gyear %>%
  select(year, q51) %>%
  group_by(year) %>%
  drop_na(q51) %>%
  summarise(avg = mean(q51))
myearq51
```

```{r}
library(tidyverse)

myearq55 <- grouped_gyear %>%
  select(year, q55) %>%
  group_by(year) %>%
  drop_na(q55) %>%
  summarise(avg = mean(q55))
myearq55
```

```{r}
library(tidyverse)

myearq56 <- grouped_gyear %>%
  select(year, q56) %>%
  group_by(year) %>%
  drop_na(q56) %>%
  summarise(avg = mean(q56))
myearq56
```

```{r}
library(tidyverse)

myearq57 <- grouped_gyear %>%
  select(year, q57) %>%
  group_by(year) %>%
  drop_na(q57) %>%
  summarise(avg = mean(q57))
myearq57
```

```{r}
library(tidyverse)

myearq58 <- grouped_gyear %>%
  select(year, q58) %>%
  group_by(year) %>%
  drop_na(q58) %>%
  summarise(avg = mean(q58))
myearq58
```

```{r}
library(tidyverse)

myearq59 <- grouped_gyear %>%
  select(year, q59) %>%
  group_by(year) %>%
  drop_na(q59) %>%
  summarise(avg = mean(q59))
myearq59
```

```{r}
library(tidyverse)

myearq60 <- grouped_gyear %>%
  select(year, q60) %>%
  group_by(year) %>%
  drop_na(q60) %>%
  summarise(avg = mean(q60))
myearq60
```

```{r}
library(tidyverse)

myearq61 <- grouped_gyear %>%
  select(year, q61) %>%
  group_by(year) %>%
  drop_na(q61) %>%
  summarise(avg = mean(q61))
myearq61
```

```{r}
library(tidyverse)

myearq62 <- grouped_gyear %>%
  select(year, q62) %>%
  group_by(year) %>%
  drop_na(q62) %>%
  summarise(avg = mean(q62))
myearq62
```

```{r}
library(tidyverse)

myearq63 <- grouped_gyear %>%
  select(year, q63) %>%
  group_by(year) %>%
  drop_na(q63) %>%
  summarise(avg = mean(q63))
myearq63
```

```{r}
library(tidyverse)

myearq64 <- grouped_gyear %>%
  select(year, q64) %>%
  group_by(year) %>%
  drop_na(q64) %>%
  summarise(avg = mean(q64))
myearq64
```

```{r}
library(tidyverse)

myearq67 <- grouped_gyear %>%
  select(year, q67) %>%
  group_by(year) %>%
  drop_na(q67) %>%
  summarise(avg = mean(q67))
myearq67
```

```{r}
library(tidyverse)

myearq68 <- grouped_gyear %>%
  select(year, q68) %>%
  group_by(year) %>%
  drop_na(q68) %>%
  summarise(avg = mean(q68))
myearq68
```

```{r}
library(tidyverse)

myearq69 <- grouped_gyear %>%
  select(year, q69) %>%
  group_by(year) %>%
  drop_na(q69) %>%
  summarise(avg = mean(q69))
myearq69
```

```{r}
library(tidyverse)

myearq70 <- grouped_gyear %>%
  select(year, q70) %>%
  group_by(year) %>%
  drop_na(q70) %>%
  summarise(avg = mean(q70))
myearq70
```

```{r}
library(tidyverse)

myearq83 <- grouped_gyear %>%
  select(year, q83) %>%
  group_by(year) %>%
  drop_na(q83) %>%
  summarise(avg = mean(q83))
myearq83
```

# Graph means across years for PCA identified questions.
```{r}
library(ggplot2)
library(dplyr) 
gq8 <-ggplot(myearq8, aes(x = year, y = avg)) + 
  geom_line(aes(color = "blue", linetype = "solid")) + 
  geom_point(aes(color = "blue")) +
  scale_x_continuous(breaks = c(1991, 1993, 1995, 1997, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 2017, 2019))
gq8 <- gq8 + labs(title = "How often do you wear a seatbelt when riding in a car driven by someone else?", x = "Year", y = "Average 4.0 = Most of the Time 5.0 = Always") +
  theme(legend.position = 'none')
gq8
```

