---
title: "Stat 850 Project Description"
author: "Jonah Garbin, Karen Alexander, Jay Jeffries"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(digits = 3)
```


```{r load-sas-libraries, echo = F, message = F, warning = F}
saspath <- "C:\\Program Files\\SASHome\\SASFoundation\\9.4\\sas.exe" # Your SAS path goes here
sasopts <- "-nosplash -ls 75"

# Please leave this in so that I can test it!
# saspath <- "/usr/local/SASHome/SASFoundation/9.4/bin/sas_en" 
# sasopts <- "-ls 75"
knitr::opts_chunk$set(
  engine.path=list(sas=saspath, saslog=saspath, sashtml=saspath),
  engine.opts=list(sas=sasopts, saslog=sasopts, sashtml=sasopts),
  comment=NA)
library(SASmarkdown)
```

## Instructions}

Each member of your team should modify this document in some way and push their modifications to the repository in a separate commit. This will ensure that you have set your repository up in a way that ensures all group members are working with the same repository. 

Note that you must compile your readme (this document) for it to properly display as part of your github repository. 

Once you have received feedback on your project proposal (via Canvas) you may alter this README so that it describes your final project instead of the project proposal. 

## Data Set
Provide a link to your dataset here, and a description of the variables in the data set.

CDC Youth Risk Behavior Survey (YRBS) website:
https://www.cdc.gov/healthyyouth/data/yrbs/data.htm

Seven kinds of variables are included in the combined datasets:
• Sample variables (identify records by survey, site type, and survey year)
• Demographic variables (self-reported demographic characteristics such as age, sex, grade, race)
• Body mass index (BMI) variables (body mass index %, an obesity indicator, and overweight indicator)
• Sexual minority variables (sexual identity and the sex of sexual contact with others)
• 2019 questionnaire variables 
• 2019 supplemental variables (dichotomous variables of broad interest; e.g. smoking, vaping, etc.)
• Selected additional risk behavior optional question variables (dichotomized risk behavior variables; e.g. driving under influence)

The dataset itself comes from the Center for Disease Control that has surveyed high school students, since 1991, to identify teen drug use, sex, eating habits, and more. The most recent year of results is 2019, and this dataset stretches across every state of the United States from high schools who have given the CDC permission for use.

The dataset is split as follows:
-The complete national dataset
-The dataset from states starting from A to M
-The dataset from states starting from N to Z
-District data

Additional background information (p. 2-5) as well as the codebook (p. 5-16) are available here:
(https://www.cdc.gov/healthyyouth/data/yrbs/pdf/2019/2019_YRBS_SADC_Documentation.pdf) 

Here is the original source for our data: (https://www.cdc.gov/healthyyouth/data/yrbs/data.htm)
We have combined the datasets from states starting from A to M and starting with N to Z in order to get all state data in one file (for analysis). Unfortunately, GitHub does not allow us to upload all 147.6 MB of data. Though, we will still be able to analyze the data here in R. 


## Potential Topics to Explore Using the Data Set
1. The timeline trends of cigar/cigarette smoking leading into vaping device smoking
2. Potential change in relationships between diet/exercise and obesity
3. How sexual identity changes with grade or age
4. Trends in opening up about sexual identity through the timeline (e.g. comparing early sexual identities in 1991 to modern sexual identities of 2019)
5. Trends in various drug use (heroin, cocaine, meth, marijuana)
6. Interaction of drug use and physical activity
7. Interactions of drug use and homelessness
8. Interaction of drug use and NOT seeing a dentist
9. Comparison of sexual identity by state/region of the U.S.
10. Comparison of obesity by state/region of the U.S.
(These potential topics are subject to change with the fluidity of our interests)

## Group Members
Karen Alexander, Jonah Garbin, and Jay Jeffries

#Outline of Analyses by Group Members
Jonah: Plans to conduct Multivariate Analyses (PCA, Cluster, and Discriminant Functions) to narrow the data, and understand how individuals aross years group together
Jay: Plans to conduct Regression Analyses (Linear Multiple, and Binary Logistic) to consider significant predictors of potential topics listed above
Karen: Plans to conduct logitudinal analyess (With Subjects ANCOVAs/regression) to consider changes in potential topics listed above over time
We then plan to do one large regresion involving all three parts: A regression of a topic above predicted by time and the componenets from the PCA

# The following code was used to edit the original data set to a smaller size as it was too large for github. 
<<<<<<< HEAD


All of this code was done in SAS as SAS could handle the file size.
```{SAS Markdown}
libname stat850 'C:\Users\kglue\Desktop\My SAS Files\9.4\STATS850\';
filename Youthrs 'C:\Users\kglue\Desktop\My SAS Files\9.4\STATS850\Youthrs.csv';

proc import datafile = 'C:\Users\kglue\Desktop\My SAS Files\9.4\STATS850\Youthrs.csv'
out=stat850.yrs
dbms = csv
replace;
getnames = yes;
PROC PRINT data = stat850.yrs (obs = 10);
RUN;

*/Delete all variables that begin with qn as they were only given in 2019 or are dichotomized versions of other variables/*
data stat850.yrs1; set stat850.yrs;
drop qn8 qn9 qn10 qn11 qn12 qn13 qn14 qn15 qn16 qn17 qn18 qn19 qn20;
run;
proc print data = stat850.yrs1 (obs = 10);
run;

*/It worked for one qn8, now trying to get it to work for rest of qn data/*
data stat850.yrs2; set stat850.yrs1;
drop qn9 qn10 qn11 qn12 qn13 qn14 qn15 qn16 qn17 qn18 qn19 qn20;
run;
proc print data = stat850.yrs2 (obs = 10);
run;

*/Okay, piece by piece I will remove qn21 - qn30 data/*
data stat850.yrs3; set stat850.yrs2;
drop qn21 qn22 qn23 qn24 qn25 qn26 qn27 qn28 qn29 qn30;
run;
proc print data = stat850.yrs3 (obs = 10);
run;
data stat850.yrs4; set stat850.yrs3;
drop qn31 qn32 qn33 qn34 qn35 qn36 qn37 qn38 qn39 qn40;
run;
proc print data = stat850.yrs4 (obs = 10);
run;
data stat850.yrs5; set stat850.yrs4;
drop qn41 qn42 qn43 qn44 qn45 qn46 qn47 qn48 qn49 qn0;
run;
proc print data = stat850.yrs5 (obs = 10);
run;
data stat850.yrs6; set stat850.yrs5;
drop qn51 qn52 qn53 qn54 qn55 qn56 qn57 qn58 qn59 qn60;
run;
proc print data = stat850.yrs6 (obs = 10);
run;
*/There is no qn65 and qn66/*
data stat850.yrs7; set stat850.yrs6;
drop qn61 qn62 qn63 qn64 qn67 qn68 qn69 qn70;
run;
proc print data = stat850.yrs7 (obs = 10);
run;
data stat850.yrs8; set stat850.yrs7;
drop qn71 qn72 qn73 qn74 qn75 qn76 qn77 qn78 qn79 qn80;
run;
proc print data = stat850.yrs8 (obs = 10);
run;
data stat850.yrs9; set stat850.yrs8;
drop qn81 qn82 qn83 qn84 qn85 qn86 qn87 qn88 qn89;
run;
proc print data = stat850.yrs9 (obs = 10);
run;

*/Now getting rid of every column that begins with qn followed by a word as those were only administered in 2019?8
data stat850.yrs10; set stat850.yrs9;
drop qnfrcig qndaycig qnfrevp qndayevp qnfrcgr qndaycgr;
run;
proc print data = stat850.yrs10 (obs = 10);
run;
data stat850.yrs11; set stat850.yrs10;
drop qntb2 qntb3 qntb4 qntb5 qniudimp qnothhpl;
run;
proc print data = stat850.yrs11 (obs = 10);
run;
data stat850.yrs12; set stat850.yrs11;
drop qndualbc qnbcnone qnobese qnowt qnfr0 qnfr1 qnfr2;
run;
proc print data = stat850.yrs12 (obs = 10);
run;
data stat850.yrs13; set stat850.yrs12;
drop qnveg0 qnveg1 qnveg2 qnveg3 qnsoda1 qnsoda2 qnfr2;
run;
proc print data = stat850.yrs13 (obs = 10);
run;
data stat850.yrs14; set stat850.yrs13;
drop qnmilk1 qnmilk3 qnbk7day qnpa0day qnpa7day qndlype qnnodnt;
run;
proc print data = stat850.yrs14 (obs = 10);
run;

*/There are still more qn categories that need removed/*
*/These are dichotomized answers which we already have without dichotomization/*
data stat850.yrs15; set stat850.yrs14;
drop qnbikehelmet qndrivemarijuana qncelldriving qnpropertydamage qnbullyweight qnbullygender qnbullygay;
run;
proc print data = stat850.yrs15 (obs = 10);
run;
data stat850.yrs16; set stat850.yrs15;
drop qnchokeself qncigschool qnchewtobschool qnalcoholschool qntypealcohol2 qnhowmarijuana qnmarijuanaschool;
run;
proc print data = stat850.yrs16 (obs = 10);
run;
data stat850.yrs17; set stat850.yrs16;
drop qncurrentopioid qncurrentcocaine qncurrentheroin qncurrentmeth qnhallucdrug qnprescription30d qnillict;
run;
proc print data = stat850.yrs17 (obs = 10);
run;
data stat850.yrs18; set stat850.yrs17;
drop qngenderexp qntaughtHIV qntaughtsexed qntaughtstd qntaughtcondom qntaughtbc qndietpop;
run;
proc print data = stat850.yrs18 (obs = 10);
run;
data stat850.yrs19; set stat850.yrs18;
drop qncoffeetea qnsportsdrink qnspdrk1 qnspdrk2 qnenergydrink qnsugardrink qnwater;
run;
proc print data = stat850.yrs19 (obs = 10);
run;
data stat850.yrs20; set stat850.yrs19;
drop qnwater1 qnwater2 qnwater3 qnfastfood qnfoodallergy qnwenthungry qnmusclestrength;
run;
proc print data = stat850.yrs20 (obs = 10);
run;
data stat850.yrs21; set stat850.yrs20;
drop qnsunscreenuse qnindoortanning qnsunburn qnconcentrating qncurrentasthma qnwheresleep qnspeakenglish;
run;
proc print data = stat850.yrs21 (obs = 10);
run;
data stat850.yrs22; set stat850.yrs21;
drop qntransgender;
run;
proc print data = stat850.yrs22 (obs = 10);
run;

*/First several columns are meaningless, they all state the same thing./*
*/That this study was in the US and this is national level data/*
*/Removing it to lower size of file./*
data stat850.yrs23; set stat850.yrs22;
drop sitecode sitename sitetypenum;
run;
proc print data = stat850.yrs23 (obs = 10);
run;

*/Export yrs23 to a csv file/*
proc export data = stat850.yrs23
	outfile = "C:\Users\kglue\Desktop\My SAS Files\9.4\STATS850\yrs24.csv"
	dbms = csv replace;
run;
=======
All of this code was done in SAS as SAS could handle the file size. 

> I've moved the code to cleaning_code.sas, which you can then reference. Please describe how you got the original data (e.g. downloaded from XXX site using YYY options) and then you can say "A script for reducing the data to the variables of interest is provided in cleaning_code.sas". The resulting data file, yrs24.csv, contains the following: <describe what you left in>

>>>>>>> f30373f7aefed20de6efb7545dd89ad84b99457c
The original code was downloaded from the following website: https://www.cdc.gov/healthyyouth/data/yrbs/data.htm
The combined Youth Risk Behavior Survey (YRBS) data was downloaded as .... The Combined Data Set's User Guide ;and the Software Analysis for YRBS were downloaded as pdfs. 

A script for reducing the data to variables of interest is provided in cleaning_code.sas. The resulting data file, yrs24,csv, contains national survey results from 1991 - 2019. The survey is conducted in odd numbered years and the national results in the file include weighted results as a complex sampling method is used. 

yrs24.csv was found to contain 22 variables with no data. Those variables were removed and the new data set was named yrisk. yrisk contains the year in which the question was asked, the sampling stratum, corresponding stratum weight and the linking variable. What is remaining in our file is data by state. Not all 50 states are included each year as inclusion of data from a state is dependent upon the state sharing the survey results with the CDC. The survey includes student demographics such as age, sex, grade, and race. Because of the changes in how race has been recorded there were originally four options for race and since 2007 there have been seven options, see https://www.cdc.gov/healthyyouth/data/yrbs/pdf/2019/2019_National_YRBS_Data_Users_Guide.pdf. 

The survey questions can be somewhat categorized into the following categories: height, weight, BMI, driving habits (q8 - q11), weapon carrying (q13 - q16), fighting (q17 - q18), dating violence (q19 - q22), bullying (q23 - q24), suicide (q25 - q29), tobacco use (q30 - q39), alcohol use (q40 - q44), marijuana use (q45 - q48), drug use (q49 - q57), sexual behaviors (q58 - q64), weight (q67 - q68), eating and drinking habits (q69 - q77), physical activity (q78 - q83), health care (q84 - q87), sleeping habits (q88), and grades in school (q89). These questions are then followed by questions that are not categorized, but represent risk. For example, qbikehelmet, asks about the frequency in which a youth wears a bike helmet when riding a bike. 

```{r}
yrisk <- readr::read_csv("yrs24.csv", guess_max = 400000)
```
```{r}
skimr::skim(yrisk)
```

# Delete the 22 variables with no data.
```{r}
yrisk <- subset(yrisk, select = -c(qcelldriving, qbullyweight, qbullygender, qbullygay, qchokeself, qtypealcohol2, qcurrentheroin, qcurrentmeth, qprescription30d, qgenderexp, qtaughtsexed, qtaughtstd, qtaughtcondom, qtaughtbc, qdietpop, qcoffeetea, qenergydrink, qsugardrink, qfastfood, qwenthungry, qwheresleep, qtransgender))
# You might be able to use e.g. colSums(!is.na(.)) to pick these columns out automatically :)
```

# Now rerun skimr with new dataset.
```{r}
skimr::skim(yrisk)
```

# Need to download survey package due to complex sampling used in survey.
```{r}
install.packages("survey")
library(survey)
```

```{r}
skimr::skim(yrisk)
```

# Want to graph output by year to see any longitudinal trends. 
```{r}
yrbsdes <- svydesign(id=~PSU, weight=~weight, strata=~stratum,data=yrisk, nest=TRUE)
```

#Summarize Data by Question and Year
```{r}
library(dplyr)
gyear <- yrisk %>%
  mutate(year) %>%
  select(year, age, sex, grade, race4, race7, stheight, stweight, bmi, bmipct, q8, q9, q10)
```

```{r}
grouped_gyear <- gyear %>%
  group_by(year)
grouped_gyear
```
  
```{r}
install.packages("psych")
library(psych)
library(tidyverse)

myfun <- function(value) {
  x <- na.omit(value)
  tibble::tibble(mean = mean(x), min = min(x), max = max(x), sd = sd(x))
}

tmp <- grouped_gyear %>%
  tidyr::pivot_longer(cols = matches("q")) %>%
  filter(year == 1991, name == "q8")

grouped_gyear %>%
  tidyr::pivot_longer(cols = matches("q")) %>%
  dplyr::group_by(year, name) %>%
  dplyr::mutate(stat = purrr::map(value, myfun))

describe.by(grouped_gyear, group=year, mean, SD, median, min, max)
```

gyear <-  yrisk %>%
  group_by(year)
  summarise(avg = mean(q8:q86))
```

```{R}
library(ggplot2)
library(dplyr)
ggplot(yrisk, aes(x = q8, y = year)) +
  geom_boxplot(color = "blue")
```             

