---
title: "Stat 850 Project Description"
author: "Jonah Garbin, Karen Alexander, Jay Jeffries"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(digits = 3)
```


```{r load-sas-libraries, echo = F, message = F, warning = F}
saspath <- "C:\\Program Files\\SASHome\\SASFoundation\\9.4\\sas.exe" # Your SAS path goes here
sasopts <- "-nosplash -ls 75"

# Please leave this in so that I can test it!
# saspath <- "/usr/local/SASHome/SASFoundation/9.4/bin/sas_en" 
# sasopts <- "-ls 75"
knitr::opts_chunk$set(
  engine.path=list(sas=saspath, saslog=saspath, sashtml=saspath),
  engine.opts=list(sas=sasopts, saslog=sasopts, sashtml=sasopts),
  comment=NA)
library(SASmarkdown)
```

## Instructions}

Each member of your team should modify this document in some way and push their modifications to the repository in a separate commit. This will ensure that you have set your repository up in a way that ensures all group members are working with the same repository. 

Note that you must compile your readme (this document) for it to properly display as part of your github repository. 

Once you have received feedback on your project proposal (via Canvas) you may alter this README so that it describes your final project instead of the project proposal. 

## Data Set
Provide a link to your dataset here, and a description of the variables in the data set.

CDC Youth Risk Behavior Survey (YRBS) website:
https://www.cdc.gov/healthyyouth/data/yrbs/data.htm

Seven kinds of variables are included in the combined datasets:
• Sample variables (identify records by survey, site type, and survey year)
• Demographic variables (self-reported demographic characteristics such as age, sex, grade, race)
• Body mass index (BMI) variables (body mass index %, an obesity indicator, and overweight indicator)
• Sexual minority variables (sexual identity and the sex of sexual contact with others)
• 2019 questionnaire variables 
• 2019 supplemental variables (dichotomous variables of broad interest; e.g. smoking, vaping, etc.)
• Selected additional risk behavior optional question variables (dichotomized risk behavior variables; e.g. driving under influence)

The dataset itself comes from the Center for Disease Control that has surveyed high school students, since 1991, to identify teen drug use, sex, eating habits, and more. The most recent year of results is 2019, and this dataset stretches across every state of the United States from high schools who have given the CDC permission for use.

The dataset is split as follows:
-The complete national dataset
-The dataset from states starting from A to M
-The dataset from states starting from N to Z
-District data

Additional background information (p. 2-5) as well as the codebook (p. 5-16) are available here:
(https://www.cdc.gov/healthyyouth/data/yrbs/pdf/2019/2019_YRBS_SADC_Documentation.pdf) 

Here is the original source for our data: (https://www.cdc.gov/healthyyouth/data/yrbs/data.htm)
We have combined the datasets from states starting from A to M and starting with N to Z in order to get all state data in one file (for analysis). Unfortunately, GitHub does not allow us to upload all 147.6 MB of data. Though, we will still be able to analyze the data here in R. 


## Potential Topics to Explore Using the Data Set
1. The timeline trends of cigar/cigarette smoking leading into vaping device smoking
2. Potential change in relationships between diet/exercise and obesity
3. How sexual identity changes with grade or age
4. Trends in opening up about sexual identity through the timeline (e.g. comparing early sexual identities in 1991 to modern sexual identities of 2019)
5. Trends in various drug use (heroin, cocaine, meth, marijuana)
6. Interaction of drug use and physical activity
7. Interactions of drug use and homelessness
8. Interaction of drug use and NOT seeing a dentist
9. Comparison of sexual identity by state/region of the U.S.
10. Comparison of obesity by state/region of the U.S.
(These potential topics are subject to change with the fluidity of our interests)

## Group Members
Karen Alexander, Jonah Garbin, and Jay Jeffries

#Outline of Analyses by Group Members
Jonah: Plans to conduct Multivariate Analyses (PCA, Cluster, and Discriminant Functions) to narrow the data, and understand how individuals aross years group together
Jay: Plans to conduct Regression Analyses (Linear Multiple, and Binary Logistic) to consider significant predictors of potential topics listed above
Karen: Plans to conduct logitudinal analyess (With Subjects ANCOVAs/regression) to consider changes in potential topics listed above over time
We then plan to do one large regresion involving all three parts: A regression of a topic above predicted by time and the componenets from the PCA

# The following code was used to edit the original data set to a smaller size as it was too large for github. 
All of this code was done in SAS as SAS could handle the file size. 

> I've moved the code to cleaning_code.sas, which you can then reference. Please describe how you got the original data (e.g. downloaded from XXX site using YYY options) and then you can say "A script for reducing the data to the variables of interest is provided in cleaning_code.sas". The resulting data file, yrs24.csv, contains the following: <describe what you left in>


```{r}
yrisk <- readr::read_csv("yrs24.csv", guess_max = 400000)
```
```{r}
skimr::skim(yrisk)
```

# Delete the 22 variables with no data.
```{r}
yrisk <- subset(yrisk, select = -c(qcelldriving, qbullyweight, qbullygender, qbullygay, qchokeself, qtypealcohol2, qcurrentheroin, qcurrentmeth, qprescription30d, qgenderexp, qtaughtsexed, qtaughtstd, qtaughtcondom, qtaughtbc, qdietpop, qcoffeetea, qenergydrink, qsugardrink, qfastfood, qwenthungry, qwheresleep, qtransgender))
# You might be able to use e.g. colSums(!is.na(.)) to pick these columns out automatically :)
```

# Now rerun skimr with new dataset.
```{r}
skimr::skim(yrisk)
```
